### 2018-03-20 CO749-GraphColoring/assignments/A4.tex
problem 4

	Let $G$ be a planar graph having girth at least five and let $S$ be a subgraph of $G$. If $G$ is $S$-critical for $3$-coloring then $|V(G)| \leq c|V(S)|$.
\end{theorem}
\begin{proof}

\end{proof}
\end{document}

>>>>

	Let $G$ be a planar graph having girth at least five and let $S$ be a subgraph of $G$. If $G$ is $S$-critical for $3$-coloring then $|V(G)| \leq c|V(S)|$.
	For $\delta_0$ as in Theorem \ref{th:four}, choose $\delta = \min\{\delta_0, \frac{1}{3}\}$, and let $\epsilon$ be the corresponding value in Theorem \ref{th:four}. Choose $c = (1+\frac{5}{\epsilon}) > 0$. Define for any graph $G$
	$$d(G) = 3e(G) - (5+\epsilon)v(G)$$
	and $d(G|S) = d(G) - d(S)$ for subgraph $S$ of $G$. Similar to the previous problem, it's clear that $d(G|S)$ satisfies the ``chain rule", that is if $S\subseteq H \subseteq G$ then
	$$d(G|S) = d(G|H) + d(H|S).$$
	\paragraph{Main Claim}
	We claim for any $G$ and $S$ as in the statement of the Theorem, $d(G|S) \geq 0$.
	Suppose for a contradiction this claim is false. Let $(G,S)$ be a counterexample minimizing $v(G|S)$. That is $G$ is a girth at least $5$ planar graph and $G$ is $S$-critical for $3$-coloring. By minimiamlity, $G$ is connected.
	$G$ has at least two vertices not in $S$ or at least $3$ edges not in $S$ (otherwise, it is easy to extend any $3$-coloring of $S$ to $G$). Let $e \in \delta(S)$. Then $S+e$ is a strict subgraph of $G$. Let $\phi$ be a $3$-coloring of $S$ which extends to $S+e$ but not to $G$. Let $H$ be the smallest subgraph of $G$ containing $S$ such that $\phi$ does not extend to $H$. So $H$ contains $S+e$, i.e. $H$ contains $S$ as a strict subgraph.
	We proceed by case distinction. Either $H=G$ or $H\subset G$. We first consider the case $H \subset G$. By minimality of $H$, every strict subgraph of $H$ can be colored by an extension of $\phi$. hence $H$ is $S$-critical. Since $H$ contains $S$, $G$ is $H$-critical. So by minimality, neither $(G,H)$ nor $(H,S)$ are counterexamples to the main claim. Thus we observe
	$$d(G|S) = d(G|H) + d(H|S) \geq 0 + 0 = 0,$$
	The more complex case is when $H=G$. In this case we see $G$ is $\phi$-critical as the minimality of our choice of $H$ implies every strict subgraph of $G$ containing $S$ is colorable with an extension of $\phi$. Let $G_S[\phi]$ denote the graph obtained from $G$ by identifying vertices of $S$ according to their color class with respect to $\phi$. Let $x_1, x_2, x_3$ be vertices of $G_S[\phi]$ corresponding to the respective color classes of $\phi$. If necessary we add edges between each $x_i, x_j$ so they form a $K_3$.
	Now we claim $G_S[\phi]$ is not $3$-colorable. Otherwise we can permute a $3$-coloring of $G_S[\phi]$ so it assigns color $i$ to $x_i$. This coloring then corresponds to an extension of $\phi$ which $3$-colors $G$ (by assigning $i$ each vertex of $G$ in $x_i$). Hence $G_S[\phi]$ contains a $4$-critical subgraph $W$. Let $S' = S \cup W - X$, where $X = W \cap \{x_1,x_2,x_3\}$.
	We claim $W$ contains $G_S[\phi] - X$. Otherwise we can extend $\phi$ to a $3$-coloring of $W$. Therefore $S' = G$. Now if we define a potential function $p(G) = -d(G)$ we can apply the Potential Extension Lemma to see
	$$p(G) = p(S') \leq p(S) + p(W) - p(X)$$
	But then rearranging we see that
	$$d(G|S) \geq p(X) + d(W).$$
	Computing the potential of $X$ we see that either $p(X) = 5+\epsilon$, $p(X) = 7 + 2\epsilon$, or $p(X) = 6+3\epsilon$ depending on whether $|X| = 1, 2,$ or $3$ respectively. So $p(X) \geq 5+\epsilon$.
	Applying Theorem \ref{th:four} to $4$-critical graph $W$ we see that
	$$d(W) \geq -2 - \delta T(W)$$
	Since $G$ is girth $5$, $T(W) \leq 3$ as the only cycles of length smaller than $5$ in $W$ intersect $\{x_1,x_2,x_3\}$ so there are at most $3$ vertex-disjoint such cycles. Hence
	$$d(W) \geq -2 - 3\delta.$$
	Thus substituting our lower bounds on $d(W)$ and $p(X)$ into the bound for $d(G|S)$ we see that
	$$d(G|S) \geq 5+\epsilon -2-3\delta \geq 1 - 3\delta \geq 0$$
	where the last inequality follows since $\delta \leq \frac{1}{3}$.
	\paragraph{Finishing the proof}
	Suppose that $G$ is a girth at least $5$ planar graph and $G$ is $S$-critical for $3$-coloring. Now we can use the Claim that $d(G|S) \geq 0$ to achieve the desired bound $|V(G)| \leq c|V(S)|$. Observe that
	$$0 \leq d(G|S) = 3e(G|S) -5v(G|S) -\epsilon v(G|S)$$
	Since $G$ is planar, girth at least $5$ we have $3e(G) -5v(G) \leq -10$. Substituting into the above yields
	$$0 \leq -10 -3e(S) +5v(S) - \epsilon v(G|S).$$ Therefore
	$$\epsilon v(G|S) \leq 5v(S)$$
	and so 
	$$v(G) \leq (1+\frac{5}{\epsilon}) v(S) = c v(S),$$
\end{document}
